---
title: "Santander Bikepoint Analysis"
output: html_document
author: "David Foster"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style>
.html-widget {
    margin: auto;
}
</style>

In this code I have designed two interactive maps which could serve as a basis for assisting in the decision of where would be best to place new Santander bike docking stations in order to benefit public health.

```{r, message = FALSE}
library(dplyr)        # data manipulation
library(data.table)   # data manipulation
library(jsonlite)     # handling json
library(httr)         # handling API requests
library(fingertipsR)  # R wrapper for public health data repository
library(sf)           # for use in handling spatial data
library(leaflet)      # map creation
library(janitor)      # data cleaning
library(RColorBrewer) # colour palettes
```

### Data sources

I'm using a number of data sources which will be accessed in the below chunk:

* The Bikepoint API:
  + Used to access the coordinates of all Santander bike docking stations in London.
  + Source: https://api.tfl.gov.uk/BikePoint/
* Deprivation data:
  + Provides information on a variety of different deprivation indices.
  + Source: https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019
* Local Authority Area & Lower Layer Super Output Area shapefiles and look-up table.
  + Used to create maps of different types of subregion of London
  + Source: https://geoportal.statistics.gov.uk/
* Public Health England profiles:
  + Provides data on a variety of public health outcomes
  + Source: https://fingertips.phe.org.uk/

```{r data_access}

# access bikepoint API
BikesAPI <- "https://api.tfl.gov.uk/BikePoint/"
get_bikes <- content(GET(BikesAPI), "text")

# convert from JSON and select relevant fields
bikes_JSON <- fromJSON(get_bikes, flatten = TRUE)
df_bikes <- as.data.frame(bikes_JSON) %>%
  select(c(id, commonName, lat, lon))

# look-up table has been accessed and downloaded from the above source
# this is just being used as it contains region data which can be joined
# to other data sources to filter countrywide data to London only
lsoa_lkp <- fread("data/lsoa_lookup.csv") %>%
  # clean headings
  clean_names() %>% 
  # filter for London and select subregions
  filter(rgn11nm == "London") %>%
  select(c(lsoa11nm, lsoa11cd, lad11nm, lad11cd))

# IMD data has been accessed and downloaded from the above source
imd <- fread("data/imd_2019.csv") %>%
  # clean headings
  clean_names() %>% 
  # inner join to lookup table and select columns
  rename("lsoa11cd" = "lsoa_code_2011") %>%
  inner_join(lsoa_lkp, by = "lsoa11cd") %>%
  select(c(lsoa11nm, lsoa11cd, lad11nm, lad11cd, index_of_multiple_deprivation_imd_rank,
           index_of_multiple_deprivation_imd_decile))

# download LSOA shapefile for mapping
lsoa_shp <- read_sf(dsn = "shapefiles",
                    layer = "LSOA_2011_EW_BGC_V3") %>%
  st_transform('+proj=longlat +datum=WGS84') %>% 
  clean_names() %>%
  # join to IMD data and limit regions to London
  inner_join(imd, by = "lsoa11cd")

# download Local Authority Area shapefile for mapping
laua_shp <- read_sf(dsn = "shapefiles",
                    layer = "Local_Authority_Districts_December_2011_GB_BFE") %>%
  st_transform('+proj=longlat +datum=WGS84') %>% 
  clean_names() %>%
  # filter by local authority area
  filter(lad11cd %in% imd$lad11cd)


# set parameters for the required indicator to download Public Health Data
# in this case I'm choosing adult physical activity levels and have looked these
# codes up previously, and the area chosen is Local Authority
indid  <- 93015
areaid <- 202

# access data and filter by region and most recent time period
df_exercise <- fingertips_data(IndicatorID = indid, AreaTypeID = 202) %>%
  filter(AreaCode %in% lsoa_lkp$lad11cd, Timeperiod == "2021/22") %>%
  rename(lad11cd = AreaCode)


# add exercise data to the Local Authority shape files
laua_shp <- laua_shp %>%
  inner_join(df_exercise, by="lad11cd")

```

### Map of docking stations and Index of Multiple deprivation deciles

The next stage is to create the maps themselves. My idea here is to have two maps. The first will show the bike docking stations imposed over an Index of Multiple Deprivation map of London. This data provides scores at LSOA level for various indices of deprivation, from health deprivation to access to green space. The 'multiple deprivation' index is an amalgamation of these separate indices which provides an overall score based on the level of deprivation in the area.

Whilst this is not a specific of an indicator as, say, childhood obesity or adult exercise levels, IMD scores are known to correlate with public health, and could form a good basis of an indicator for where new bike docking stations could be placed in order to benefit the population as a whole.


```{r imd_map}

# set colour palette for LSOA shapes
imd_cols <- colorFactor(palette = "RdYlBu",
                         domain = lsoa_shp$index_of_multiple_deprivation_imd_decile)

# create map
leaflet() %>% 
  
  # set view of correct area, I've just centered on a point in central London
  # but map can be scrolled
  setView(lat = 51.49580, lng = -0.127575, zoom = 11.2) %>%
  
  # this is used to generate the base map of London
  addProviderTiles(provider = providers$CartoDB.Positron) %>% 
  
  # add polygons to represent LSOAs, colour-coded according to the IMD decile
  addPolygons(data = lsoa_shp,
              fillColor = ~imd_cols(index_of_multiple_deprivation_imd_decile),
              fillOpacity = 0.8,
              stroke = T,
              weight = 1,
              color = "grey") %>% 
  
  # add markers to represent bikepoints
  addCircleMarkers(data = df_bikes,
                   radius = 4, 
                   stroke = F, 
                   fillColor = "black",
                   fillOpacity = 0.9,
                   lat = ~lat,
                   lng = ~lon) %>%
  
  # add further imposed polygons to represent the larger subregions of the
  # local authority areas
  addPolygons(data = laua_shp,
              fillOpacity = 0,
              highlightOptions = options(color = "grey20",
                                         weight = 4),
              label = paste0("Local Authority 2011: ", laua_shp$lad11nm))


```

### Map of docking stations and adult physical activity levels

The second map imposes the bike docking stations on a traffic-light coded map of the relevant Local Authority Areas. The traffic light system represents the level of adult physical exercise in relation to other areas of London, and could serve as another indicator of where docking stations could aid public health by encouraging exercise. 


```{r physact_map}

# create basic comparative traffic light colour palette
pal <- colorFactor(palette = c("yellow", "red", "green"), 
              levels = c("Similar", "Worse", "Better"))

# create map
leaflet() %>% 

  # set view of correct area, I've just centered on a point in central London
  # but map can be scrolled  
  setView(lat = 51.49580, lng = -0.127575, zoom = 11.2) %>%
  
  # this is used to generate the base map of London  
  addProviderTiles(provider = providers$CartoDB.Positron) %>% 
  
  # add markers to represent bikepoints  
  addCircleMarkers(data = df_bikes,
                   radius = 3, 
                   stroke = F, 
                   fillColor = "black",
                   fillOpacity = 0.9,
                   lat = ~lat,
                   lng = ~lon) %>%
  
  # add shapes to represent local authority areas, highlighted to show the relative
  # levels of adult physical activity
  addPolygons(data = laua_shp,
              highlightOptions = options(color = "grey20",
                                         weight = 4),
              label = paste0("Local Authority 2011: ", laua_shp$lad11nm),
              fillColor = ~pal(ComparedtoRegionvalueorpercentiles))


```
